Text embedding models yield high resolution insights into conceptual knowledge from short multiple choice quizzes

### 2023-01-31 paper/main.tex
checking in edits to intro + results

Another, more sensitive, way of summarizing the conceptual content of the
lectures and questions is to look at \textit{variability} in how topics are
weighted over time and across different questions (Fig.~\ref{fig:topics}).
Intuitively, the variability in the expression of a given topic relates to how
much ``information''~\citep{Fish22} the lecture (or questions) reflect about
that topic. For example, suppose a given topic is weighted on heavily
throughout a lecture. That topic might be characteristic of some aspect or
property of the lecture \textit{overall} (conceptual or otherwise), but unless
the topic's weights changed in meaningful ways over time, the topic would be a
poor indicator of any \textit{specific} conceptual content in the lecture. We
therefore also compared the variance in topic weights (across time or
questions) between the lectures and questions.  [ADD BACK CORRELATION RESULTS]

\textbf{JRM STOPPED HERE...}

We used the fitted topic model to transform the text of each question in our
stimulus pool into the same embedding space as the two lectures' trajectories
(Fig.~\ref{fig:sliding-windows}C). This yielded a single 15-dimensional
coordinate (i.e., topic vector) for each question. We then computed the
variance of each topic's weight across timepoints of each lecture, and across
questions from each category (i.e., questions about lecture 1, lecture 2, and
general physics knowledge; Fig.~\ref{fig:topics}A). Visual inspection of
Figure~\ref{fig:topics}A suggests a strong correspondence between the sets of
topics expressed by each lecture and its related questions, with only limited
overlap in the topics expressed by non-matched lecture-question set pairs. To
quantify these apparent similarities and differences, we computed the
correlations between each pair of topic-weight variance vectors (i.e., vectors
containing the variance across timepoints or questions, for each topic;
Fig.~\ref{fig:topics}B).

We found that our model represented the contents of
lecture-related questions using a combination of topic dimensions (and relative
variation along those topic dimensions) that was highly similar to that of
their reference lecture (\textit{Four Fundamental Forces} (\textit{FFF})
questions vs.~lecture: Pearson's $r(13) =
0.824,~p<0.001,~95\%~\mathrm{confidence~interval~(CI)} = [0.696, 0.973]$;
\textit{Birth of Stars} (\textit{BoS}) questions vs.~lecture: $r(13) =
0.801,~p<0.001,~95\%~\mathrm{CI} = [0.539, 0.958]$) but diverged significantly
from that of the non-reference lecture (\textit{FFF} questions vs.~\textit{BoS}
lecture: $r(13) = -0.620,~p = 0.014,~95\%~\mathrm{CI} = [-0.871, -0.326]$;
\textit{BoS} questions vs.~\textit{FFF} lecture: $r(13) = -0.550,~p =
0.034,~95\%~\mathrm{CI} = [-0.803, -0.246]$). This indicated that our model
captured the conceptual contents of the lectures and quiz questions
sufficiently well to differentiate between questions relating to one lecture
versus the other.

@@

The ability to quantify how much each question is ``asking about'' the content
from each moment of the lectures could enable high-resolution insights into
participants' knowledge. Traditional approaches to estimating how much a
student ``knows'' about the content of a given lecture entail computing the
proportion of correctly answered questions. But if two students receive
identical scores on an exam, might our modeling framework help us to gain more
nuanced insights into the \textit{specific} content that each student has
mastered (or failed to master)? For example, a student who misses three
questions that were all about the same concept (e.g., concept $A$) will have
gotten the same \textit{proportion} of questions correct as another student who
missed three questions about three \textit{different} concepts (e.g., $A$, $B$,
and $C$). But if we wanted to fill in the ``gaps'' in the two students'
understandings, we might do well to focus on concept $A$ for the first student,
but to also add in materials pertaining to concepts $B$ and $C$ for the second
student.

>>>>

Another, more sensitive, way of summarizing the conceptual content of the
lectures and questions is to look at \textit{variability} in how topics are
weighted over time and across different questions (Fig.~\ref{fig:topics}).
Intuitively, the variability in the expression of a given topic relates to how
much ``information''~\citep{Fish22} the lecture (or questions) reflect about
that topic. For example, suppose a given topic is weighted on heavily
throughout a lecture. That topic might be characteristic of some aspect or
property of the lecture \textit{overall} (conceptual or otherwise), but unless
the topic's weights changed in meaningful ways over time, the topic would be a
poor indicator of any \textit{specific} conceptual content in the lecture. We
therefore also compared the variance in topic weights (across time or
questions) between the lectures and questions. The variability in topic
expression (over time and across questions) was similar for the lecture 1 video
and questions ($r(13) = 0.824,~p<0.001, CI = [0.696, 0.973]$) and the lecture 2
video and questions ($r(13) = 0.801,~p<0.001,~95\%~\mathrm{CI} = [0.539,
0.958]$). However, as reported in Figure~\ref{fig:topics}B, the variability in
topic expressions across \textit{different} videos and lecture-specific
questions (i.e., lecture 1 video versus lecture 2 questions; lecture 2 video
versus lecture 1 questions) were \textit{negatively} correlated, and neither
video's topic variability was reliably correlated with the topic variability
across general physics knowledge questions. Taken together, the analyses
reported in Figures~\ref{fig:topics} and~\topicWeights~indicate that a topic
model fit to the videos' transcripts can also reveal correspondances (at a
coarse scale) between the lectures and (held-out) questions.

@@

The ability to quantify how much each question is ``asking about'' the content
from each moment of the lectures could enable high-resolution insights into
participants' knowledge. Traditional approaches to estimating how much a
student ``knows'' about the content of a given lecture entail computing the
proportion of correctly answered questions. But if two students receive
identical scores on an exam, might our modeling framework help us to gain more
nuanced insights into the \textit{specific} content that each student has
mastered (or failed to master)? For example, a student who misses three
questions that were all about the same concept (e.g., concept $A$) will have
gotten the same \textit{proportion} of questions correct as another student who
missed three questions about three \textit{different} concepts (e.g., $A$, $B$,
and $C$). But if we wanted to fill in the ``gaps'' in the two students'
understandings, we might do well to focus on concept $A$ for the first student,
but to also add in materials pertaining to concepts $B$ and $C$ for the second
student. In other words, raw ``proportion-correct'' measures may capture
\textit{how much} a student knows, but not \textit{what} they know. We wondered
whether our modeling framework might enable us to (formally and automatically)
infer participants' knowledge at the scale of individual concepts (e.g., as
captured by a single question).
